Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-05 paper-new/intro.tex
[paper] some minor fixes

The remainder of the paper is organized as follows. In Section
\ref{sec:technical-background}, we provide technical background on synthetic
guarded domain theory and some important details about the specific variant of
SGDT we employ as our ambient language.
%
In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
calculus for gradual typing, which we also refer to as the \emph{gradually-typed
lambda calculus}. We do this to establish a common syntax that will be
interpreted into the models we subsequently construct.
%
In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
semantics of gradually typed languages with the goal of organizing our
construction of denotational models of gradual typing.
%
In Sections \ref{sec:concrete-term-model} and
\ref{sec:towards-relational-model}, we set out to instantiate the abstract
categorical semantics by constructing a concrete model in a suitable double
category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
model for the \emph{terms} of the gradually-typed lambda calculus. This involves
using the tools of SGDT to solve a domain equation to give the semantics of the
dynamic type. We explain how to \emph{globalize} this model to obtain a model in
the usual (non-guarded) dependent type theory. In the end, we obtain a
\emph{big-step semantics} for closed terms of type $\nat$:
%
\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{nat} + 1)) \] 
%
where a closed term $M$ of type $\nat$ denotes a \emph{function} from
$\mathbb{N}$ to $\sem{nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
fails to terminate in $i$ or fewer steps. This is already a significant result,
as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
%
Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
methodology of the previous section to give a denotational semantics that models
the relational theory of the gradually-typed lambda calculus in addition to the
terms. Interesting issues arise here arising from the \emph{intensional} nature
of SGDT, that is, the fact that \emph{steps are observable}. We then develop
novel solutions to address these issues. We do not complete the construction of
the concrete relational model in this section. Instead, we choose to reformulate
the definition of abstract categorical model to take into account the solutions
to the issues discussed here.

>>>>

The remainder of the paper is organized as follows. In Section
\ref{sec:technical-background}, we provide technical background on synthetic
guarded domain theory and some important details about the specific variant of
SGDT we employ as our ambient language.
%
In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
calculus for gradual typing, which we also refer to as the \emph{gradually-typed
lambda calculus}. We do this to establish a common syntax that will be
interpreted into the models we subsequently construct.
%
In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
semantics of gradually typed languages with the goal of organizing our
construction of denotational models of gradual typing.
%
In Sections \ref{sec:concrete-term-model} and
\ref{sec:towards-relational-model}, we set out to instantiate the abstract
categorical semantics by constructing a concrete model in a suitable double
category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
model for the \emph{terms} of the gradually-typed lambda calculus. This involves
using the tools of SGDT to solve a domain equation to give the semantics of the
dynamic type. We explain how to \emph{globalize} this model to obtain a model in
the usual (non-guarded) dependent type theory. In the end, we obtain a
\emph{big-step semantics} for closed terms of type $\nat$:
%
\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{\nat} + 1)) \] 
%
where a closed term $M$ of type $\nat$ denotes a \emph{function} from
$\mathbb{N}$ to $\sem{\nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
fails to terminate in $i$ or fewer steps. This is already a significant result,
as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
%
Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
methodology of the previous section to give a denotational semantics that models
the relational theory of the gradually-typed lambda calculus in addition to the
terms. Interesting issues arise here arising from the \emph{intensional} nature
of SGDT, that is, the fact that \emph{steps are observable}. We then develop
novel solutions to address these issues. We do not complete the construction of
the concrete relational model in this section. Instead, we choose to reformulate
the definition of abstract categorical model to take into account the solutions
to the issues discussed here.

### 2024-07-05 paper-new/technical-background.tex
[paper] some minor fixes

SGDT offers a synthetic approach to domain theory that allows for guarded recursion
to be expressed syntactically via a type constructor $\later : \type \to \type$ 
(pronounced ``later''). The use of a modality to express guarded recursion
was introduced by Nakano \cite{Nakano2000}.
%
Given a type $A$, the type $\later A$ represents an element of type $A$
that is available one time step later. There is an operator $\nxt : A \to\, \later A$
that ``delays'' an element available now to make it available later.
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.

>>>>

SGDT offers a synthetic approach to domain theory that allows for guarded recursion
to be expressed syntactically via a type constructor $\later \colon \type \to \type$ 
(pronounced ``later''). The use of a modality to express guarded recursion
was introduced by Nakano \cite{Nakano2000}.
%
Given a type $A$, the type $\later A$ represents an element of type $A$
that is available one time step later. There is an operator $\nxt : A \to\, \later A$
that ``delays'' an element available now to make it available later.
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.
